Contracts for unions and intersections are hard for Nickel

### 2021-06-04 paper.tex
Acquire lock for section 4
Rework beginning of section 4

Type systems featuring intersection types are hard in a fundamental way,
from their discovery, they've been studied as a form to characterize
lambda-calculus terms normalizability; which means that typechecking a
term on a system with intersection types is undecidable,
similar to how the halting problem is undecidable\todo{reference}.

Usually, when thinking about programming languages, there's an intuition
that static checks (like typechecking) are more difficult and less powerful
than dynamic checks (like testing).
For instance, it is very easy to check if,
in a given execution of a program, a variable holds an integer value; while it
may become more challenging to check if a variable will always hold an integer value,
no matter the particular execution.

As we intend to show on this paper, unions and intersections are not only
statically hard to check, but they are also dynamically hard:
\emph{they are fundamentally hard}.
In this section, we present intuitively the main challenges that arise
when attempting to develop a runtime check system for unions and intersections.

@@

When programming with dynamic checks it is assumed that lack of
failure does not mean that the program works perfectly.

For instance, imagine having a function wrapped in a contract
that checks that it maps positive numbers to positive numbers,
and using it in different ways,
like the program on Figure \ref{fig:pos-to-pos}.
Clearly, there is an error on this program, but if the first call
to \nickel{f} that gets executed is \nickel{f 10}, the error will not
be found at that point, and the first operand of the addition would be
evaluated to \nickel{3}.
If, later on, \nickel{f 5} gets executed, the contract would get violated,
since \nickel{f} gets called with a positive value, but returns a negative
one (\nickel{-2}).

@@

Even if the program performed two different applications of \nickel{f}
before finding the contract violation, the first one is not actually
needed, and executing only \nickel{f 5} is enough to find the problem.

Having a witness of size 1 for every contract failure is a really nice to have property
since it means, for example, that every time that a contract is violated,
the stack trace is enough to understand how it was violated, and to potentially
come up with a minimal example that breaks the contract.
However, when unions get introduced, this property is lost.

Consider now the program presented in Figure \ref{fig:wrong-union-function},
this program (and in particular, \nickel{f}'s contract) still presents a problem,
since \nickel{fun x => x - 7} does not map positive numbers to positive numbers,
nor does it map positive numbers to non-positive numbers, so it won't be a valid
value for the union of those contracts.

@@

As before, the program presents enough function calls to find that issue
and violate the contract,
since \nickel{f 10} is a witness of \nickel{f} failing \nickel{Positive -> NonPositive},
and \nickel{f 5} is a witness of \nickel{f} failing \nickel{Positive -> Positive}.
But, what makes this case completely different from the one presented before, is that
both function calls are needed, and, what is more, there is not a way to
use the wrapped value \nickel{f} only once that would find a violation to its contract;
at least two are needed.

>>>>

Static typing is about enforcing that some property always holds for a
potentially unknown value, no matter the execution part. A dynamic check, on the
other hand, just need to verify a property for a specific and known value, which
is usually much easier. For example, dependent function types are complex and
mostly implemented in proof assistants (although they are slowly making their
way in some mainstream languages), whereas the corresponding dependent function
contracts can be implemented in a few line of code (as is done in
Racket~\cite{}).

The gist of the present paper is to bring evidence that in the case of union and
intersection, the dynamic checking part is complex as well: \emph{unions and
intersections are fundamentally hard}.

% Teo's version
% Implementing a static type system is not an easy task.
% Type systems featuring intersection types are hard in a fundamental way,
% from their discovery, they've been studied as a form to characterize
% lambda-calculus terms normalizability; which means that typechecking a
% term on a system with intersection types is undecidable,
% similar to how the halting problem is undecidable\todo{reference}.
%
% Usually, when thinking about programming languages, there's an intuition
% that static checks (like typechecking) are more difficult and less powerful
% than dynamic checks (like testing).
% For instance, it is very easy to check if,
% in a given execution of a program, a variable holds an integer value; while it
% may become more challenging to check if a variable will always hold an integer value,
% no matter the particular execution.
%
% As we intend to show on this paper, unions and intersections are not only
% statically hard to check, but they are also dynamically hard:
% \emph{they are fundamentally hard}.
% In this section, we present intuitively the main challenges that arise
% when attempting to develop a runtime check system for unions and intersections.

@@

In a pure language with standard contracts, the failure of a function contract
can always be blamed on a single call. Figure~\ref{fig:pos-to-pos} shows a
function with a simple contract attached that ensures that both the argument and
the result of \nickel{f} are positive numbers. \nickel{f} is called twice, and
the second call violates the requirement on the codomain, leading to blame
\nickel{f}. The first call to \nickel{f} does not matter, and \nickel{5} is a
single and independent witness of the violation of the contract. The user can be
pointed to one location in practice.

There is an intuitive justification for this fact. Apart from the error
reporting part (which is crucial in practice!), the current contract system of
Nickel can be implemented purely as a library, just given a \nickel{fail}
primitive to abort the execution. In particular, applying a function contract to
\nickel{f} replaces it with a guarding function \nickel{fWrapped} that performs
the required checks. Thus, in a pure setting, the failure of \nickel{fWrapped 5}
must be independant of its environment, and in particular of the call
\nickel{fWrapped 10}.

% When programming with dynamic checks it is assumed that lack of
% failure does not mean that the program works perfectly.
%
% For instance, imagine having a function wrapped in a contract
% that checks that it maps positive numbers to positive numbers,
% and using it in different ways,
% like the program on Figure \ref{fig:pos-to-pos}.
% Clearly, there is an error on this program, but if the first call
% to \nickel{f} that gets executed is \nickel{f 10}, the error will not
% be found at that point, and the first operand of the addition would be
% evaluated to \nickel{3}.
% If, later on, \nickel{f 5} gets executed, the contract would get violated,
% since \nickel{f} gets called with a positive value, but returns a negative
% one (\nickel{-2}).

@@

% Even if the program performed two different applications of \nickel{f}
% before finding the contract violation, the first one is not actually
% needed, and executing only \nickel{f 5} is enough to find the problem.
%
% Having a witness of size 1 for every contract failure is a really nice to have property
% since it means, for example, that every time that a contract is violated,
% the stack trace is enough to understand how it was violated, and to potentially
% come up with a minimal example that breaks the contract.
% However, when unions get introduced, this property is lost.

Union contracts are a different matter. Consider the program presented in
Figure \ref{fig:wrong-union-function}. \nickel{f} is again violating the
contract, as it neither maps positive numbers to positive numbers nor positive
numbers to negative numbers.

@@

In a system with union and intersection, this contract would fail, because
\nickel{f 10} is a witness of \nickel{f} failing \nickel{Positive -> NonPositive},
and \nickel{f 5} is a witness of \nickel{f} failing \nickel{Positive -> Positive}.
But, as opposed to the previous example, removing one call or the other will make the program pass!
Indeed, the first call only fails the \nickel{Positive -> NonPositive} part of
the union,
while the second one only fails the \nickel{Positive -> NonPositive} part. Both
are needed to prove the violation of the full contract.

Even in a pure setting, unions introduce a form of effect at a distance: the
result of \nickel{f 5} is not local anymore as it depends on the previous
execution of other calls to \nickel{f}. With unions,
equation~\ref{eq:cbn-expansion} of Section~\ref{sec:optimizations} does not
hold, and the candidate example of
Figure~\ref{fig:optimizations-let-floating-ex} cannot be optimized in general.

Figure~\ref{fig:optimized-programs} below shows another example of the same
fact. A seemingly valid optimization ends up with a program that behave
differently from the original one.

\unsure{WARNING: what is below is being reworked by Yann}
